---
title: 'Additive Model - Block 1 Flexibility Simulations'
subtitle: 'Variation of p(L), p(I), and correlation'
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
editor_options:
  markdown:
    wrap: 72
---

## Introduction

This script runs Block 1 flexibility simulations for the Additive Model,
systematically varying three parameters:
- p(L): Probability of latent disorder (0.5, 0.65, 0.8)
- p(I): Probability of inflammation (0.1, 0.3, 0.5)
- c: Correlation between L and I (0.1, 0.3, 0.5)

This creates 3 × 3 × 3 = 27 scenarios. For each scenario, we run 1000
simulations and calculate heterogeneity indices (Shannon index and Hill
numbers q=0, q=1, q=2) for both inflammation groups.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Set-up

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(mvtnorm)
library(MASS)
library(knitr)
library(flextable)
library(officer)
```

```{r general_setup, echo=TRUE}
# Setting seed for reproducibility
set.seed(123)

# Combinations of binary criteria
combinations <- expand.grid(c(0,1), c(0,1), c(0,1), c(0,1), c(0,1))
```

## Helper Functions

```{r helper_functions, echo=TRUE}
# Function to ensure the values fall between 0 and 4
rescale <- function(x) {
  min_val <- min(x)
  max_val <- max(x)
  return((x - min_val) / (max_val - min_val) * 4)
}

# Function to generate correlated binary variables
generate_correlated_binary <- function(n, p1, p2, cor) {
  means <- c(0, 0)
  sigma <- matrix(c(1, cor, cor, 1), 2, 2)
  bvn <- mvrnorm(n, mu = means, Sigma = sigma)
  
  threshold_L <- qnorm(1 - p1)
  threshold_I <- qnorm(1 - p2)
  
  L <- ifelse(bvn[,1] > threshold_L, 1, 0)
  I <- ifelse(bvn[,2] > threshold_I, 1, 0)
  
  return(data.frame(L = L, I = I))
}

# Function to calculate Shannon index
calculate_shannon <- function(probabilities) {
  p <- probabilities[probabilities > 0]
  p <- p / sum(p)
  H <- -sum(p * log(p))
  return(H)
}

# Function to calculate Hill numbers
calculate_hill <- function(probabilities, q) {
  p <- probabilities[probabilities > 0]
  p <- p / sum(p)
  
  if (q == 1) {
    # Limit case: exponential of Shannon entropy
    return(exp(-sum(p * log(p))))
  } else {
    # General formula: Hill = (sum(p^q))^(1/(1-q))
    return(sum(p^q)^(1/(1-q)))
  }
}
```

## Simulation Parameters

```{r parameters, echo=TRUE}
# Number of individuals per simulation
n <- 10000

# Number of simulations per scenario
n_sims <- 100

# Parameter grid for Block 1
p_L_values <- c(0.5, 0.65, 0.8)
p_I_values <- c(0.1, 0.3, 0.5)
cor_values <- c(0.1, 0.3, 0.5)

# Create full parameter grid
parameter_grid <- expand.grid(
  p_L = p_L_values,
  p_I = p_I_values,
  cor = cor_values
)

cat("Total number of scenarios:", nrow(parameter_grid), "\n")
cat("Simulations per scenario:", n_sims, "\n")
cat("Total simulations:", nrow(parameter_grid) * n_sims, "\n")
```

## Results Storage

```{r results_setup, echo=TRUE}
# Create results dataframe with RAW NUMERIC values (not formatted strings)
# This allows flexible formatting later for different table types

results_block1 <- data.frame(
  # Parameters
  p_L = numeric(),
  p_I = numeric(),
  cor = numeric(),
  
  # Sample sizes (mean only)
  n_I0_mean = numeric(),
  n_I1_mean = numeric(),
  
  # Severity (mean only)
  severity_I0_mean = numeric(),
  severity_I1_mean = numeric(),
  
  # Shannon: mean, SD for groups; mean, CI for delta
  shannon_I0_mean = numeric(),
  shannon_I0_sd = numeric(),
  shannon_I1_mean = numeric(),
  shannon_I1_sd = numeric(),
  delta_shannon_mean = numeric(),
  delta_shannon_ci_lower = numeric(),
  delta_shannon_ci_upper = numeric(),
  
  # Hill q=0: mean, SD for groups; mean, CI for delta
  hill_q0_I0_mean = numeric(),
  hill_q0_I0_sd = numeric(),
  hill_q0_I1_mean = numeric(),
  hill_q0_I1_sd = numeric(),
  delta_hill_q0_mean = numeric(),
  delta_hill_q0_ci_lower = numeric(),
  delta_hill_q0_ci_upper = numeric(),
  
  # Hill q=1: mean, SD for groups; mean, CI for delta
  hill_q1_I0_mean = numeric(),
  hill_q1_I0_sd = numeric(),
  hill_q1_I1_mean = numeric(),
  hill_q1_I1_sd = numeric(),
  delta_hill_q1_mean = numeric(),
  delta_hill_q1_ci_lower = numeric(),
  delta_hill_q1_ci_upper = numeric(),
  
  # Hill q=2: mean, SD for groups; mean, CI for delta
  hill_q2_I0_mean = numeric(),
  hill_q2_I0_sd = numeric(),
  hill_q2_I1_mean = numeric(),
  hill_q2_I1_sd = numeric(),
  delta_hill_q2_mean = numeric(),
  delta_hill_q2_ci_lower = numeric(),
  delta_hill_q2_ci_upper = numeric(),
  
  stringsAsFactors = FALSE
)
```

## Running Simulations

```{r simulations, echo=TRUE}
# Loop through each parameter combination
for(param_idx in 1:nrow(parameter_grid)) {
  
  # Extract current parameter values
  current_p_L <- parameter_grid$p_L[param_idx]
  current_p_I <- parameter_grid$p_I[param_idx]
  current_cor <- parameter_grid$cor[param_idx]
  
  # Progress indicator
  cat("\n=== Processing scenario", param_idx, "of", nrow(parameter_grid), "===\n")
  cat("p(L) =", current_p_L, ", p(I) =", current_p_I, ", cor =", current_cor, "\n")
  
  # Vectors to store results for this scenario
  n_I0_vec <- numeric(n_sims)
  n_I1_vec <- numeric(n_sims)
  severity_I0_vec <- numeric(n_sims)
  severity_I1_vec <- numeric(n_sims)
  shannon_I0_vec <- numeric(n_sims)
  shannon_I1_vec <- numeric(n_sims)
  hill_q0_I0_vec <- numeric(n_sims)
  hill_q0_I1_vec <- numeric(n_sims)
  hill_q1_I0_vec <- numeric(n_sims)
  hill_q1_I1_vec <- numeric(n_sims)
  hill_q2_I0_vec <- numeric(n_sims)
  hill_q2_I1_vec <- numeric(n_sims)
  
  # Run simulations for this scenario
  for(sim in 1:n_sims) {
    
    # Generate correlated binary latent variables
    latent_vars <- generate_correlated_binary(n, p1 = current_p_L, p2 = current_p_I, cor = current_cor)
    L <- latent_vars$L
    I <- latent_vars$I
    
    # Simulate symptoms with additive model (fixed coefficients)
    S1 <- rescale(2.0 * L + 0.6 * I + rnorm(n, mean=2, sd=0.5))
    S2 <- rescale(1.8 * L + 0.5 * I + rnorm(n, mean=2, sd=0.5))
    M  <- rescale(1.2 * L + 0.7 * I + rnorm(n, mean=2, sd=0.8))
    W1 <- rescale(0.6 * L + 0.5 * I + rnorm(n, mean=2, sd=1))
    W2 <- rescale(0.4 * L + 0.3 * I + rnorm(n, mean=2, sd=1))
    
    # Combine into data frame
    data <- data.frame(L, I, S1, S2, W1, W2, M)
    
    # Create binary symptom indicators
    data <- data %>%
      mutate(
        S1_positive = S1 >= 2,
        S2_positive = S2 >= 2,
        W1_positive = W1 >= 2,
        W2_positive = W2 >= 2,
        M_positive = M >= 2
      )
    
    # Count criteria met
    data <- data %>%
      mutate(criteria_met = S1_positive + S2_positive + W1_positive + W2_positive + M_positive)
    
    # Filter to diagnosed individuals (≥2 criteria)
    filtered_data <- data %>% filter(criteria_met >= 2)
    
    # Calculate symptom severity
    filtered_data <- filtered_data %>%
      mutate(severity = S1 + S2 + M + W1 + W2)
    
    # Split by inflammation status
    filtered_WITH_I <- filtered_data %>% filter(I == 1)
    filtered_WITHOUT_I <- filtered_data %>% filter(I == 0)
    
    # Skip if either group is too small
    if(nrow(filtered_WITH_I) < 10 || nrow(filtered_WITHOUT_I) < 10) {
      next
    }
    
    # === Severity-based sampling for equal groups ===
    mean_severity_WITH <- mean(filtered_WITH_I$severity)
    sd_severity_WITH <- sd(filtered_WITH_I$severity)
    
    lower_bound <- mean_severity_WITH - 0.5 * sd_severity_WITH
    upper_bound <- mean_severity_WITH + 0.5 * sd_severity_WITH
    
    filtered_WITHOUT_I_eligible <- filtered_WITHOUT_I %>%
      filter(severity >= lower_bound & severity <= upper_bound)
    
    n_sample <- min(nrow(filtered_WITH_I), nrow(filtered_WITHOUT_I_eligible))
    
    if(n_sample > 5) {
      filtered_WITH_I <- filtered_WITH_I %>% sample_n(n_sample)
      filtered_WITHOUT_I <- filtered_WITHOUT_I_eligible %>% sample_n(n_sample)
    } else {
      next
    }
    
    # Track sample sizes and severity
    n_I1_vec[sim] <- nrow(filtered_WITH_I)
    n_I0_vec[sim] <- nrow(filtered_WITHOUT_I)
    severity_I1_vec[sim] <- mean(filtered_WITH_I$severity)
    severity_I0_vec[sim] <- mean(filtered_WITHOUT_I$severity)
    
    # Create combination strings
    filtered_WITH_I <- filtered_WITH_I %>%
      mutate(combination = paste0(as.integer(S1_positive),
                                   as.integer(S2_positive),
                                   as.integer(W1_positive),
                                   as.integer(W2_positive),
                                   as.integer(M_positive)))
    
    filtered_WITHOUT_I <- filtered_WITHOUT_I %>%
      mutate(combination = paste0(as.integer(S1_positive),
                                   as.integer(S2_positive),
                                   as.integer(W1_positive),
                                   as.integer(W2_positive),
                                   as.integer(M_positive)))
    
    # Count combinations and calculate proportions
    counts_WITH <- filtered_WITH_I %>%
      count(combination) %>%
      mutate(proportion = n / sum(n))
    
    counts_WITHOUT <- filtered_WITHOUT_I %>%
      count(combination) %>%
      mutate(proportion = n / sum(n))
    
    # Calculate heterogeneity indices for WITH inflammation (I=1)
    if(nrow(counts_WITH) > 0) {
      shannon_I1_vec[sim] <- calculate_shannon(counts_WITH$proportion)
      hill_q0_I1_vec[sim] <- calculate_hill(counts_WITH$proportion, q = 0)
      hill_q1_I1_vec[sim] <- calculate_hill(counts_WITH$proportion, q = 1)
      hill_q2_I1_vec[sim] <- calculate_hill(counts_WITH$proportion, q = 2)
    }
    
    # Calculate heterogeneity indices for WITHOUT inflammation (I=0)
    if(nrow(counts_WITHOUT) > 0) {
      shannon_I0_vec[sim] <- calculate_shannon(counts_WITHOUT$proportion)
      hill_q0_I0_vec[sim] <- calculate_hill(counts_WITHOUT$proportion, q = 0)
      hill_q1_I0_vec[sim] <- calculate_hill(counts_WITHOUT$proportion, q = 1)
      hill_q2_I0_vec[sim] <- calculate_hill(counts_WITHOUT$proportion, q = 2)
    }
  }
  
  # Remove zero/NA values (failed simulations)
  valid_idx <- n_I1_vec > 0 & n_I0_vec > 0
  
  if(sum(valid_idx) < 10) {
    cat("WARNING: Too few valid simulations for this scenario!\n")
    next
  }
  
  # Filter to valid simulations
  n_I0_vec <- n_I0_vec[valid_idx]
  n_I1_vec <- n_I1_vec[valid_idx]
  severity_I0_vec <- severity_I0_vec[valid_idx]
  severity_I1_vec <- severity_I1_vec[valid_idx]
  shannon_I0_vec <- shannon_I0_vec[valid_idx]
  shannon_I1_vec <- shannon_I1_vec[valid_idx]
  hill_q0_I0_vec <- hill_q0_I0_vec[valid_idx]
  hill_q0_I1_vec <- hill_q0_I1_vec[valid_idx]
  hill_q1_I0_vec <- hill_q1_I0_vec[valid_idx]
  hill_q1_I1_vec <- hill_q1_I1_vec[valid_idx]
  hill_q2_I0_vec <- hill_q2_I0_vec[valid_idx]
  hill_q2_I1_vec <- hill_q2_I1_vec[valid_idx]
  
  cat("Valid simulations:", sum(valid_idx), "\n")
  
  # Calculate deltas
  delta_shannon <- shannon_I1_vec - shannon_I0_vec
  delta_hill_q0 <- hill_q0_I1_vec - hill_q0_I0_vec
  delta_hill_q1 <- hill_q1_I1_vec - hill_q1_I0_vec
  delta_hill_q2 <- hill_q2_I1_vec - hill_q2_I0_vec
  
  # Helper function to calculate 95% CI
  calc_ci <- function(x) {
    m <- mean(x)
    se <- sd(x) / sqrt(length(x))
    return(c(lower = m - 1.96 * se, upper = m + 1.96 * se))
  }
  
  # Calculate CIs for deltas
  ci_shannon <- calc_ci(delta_shannon)
  ci_hill_q0 <- calc_ci(delta_hill_q0)
  ci_hill_q1 <- calc_ci(delta_hill_q1)
  ci_hill_q2 <- calc_ci(delta_hill_q2)
  
  # Add row to results with RAW NUMERIC values
  results_block1 <- rbind(results_block1, data.frame(
    # Parameters
    p_L = current_p_L,
    p_I = current_p_I,
    cor = current_cor,
    
    # Sample sizes (mean only)
    n_I0_mean = mean(n_I0_vec),
    n_I1_mean = mean(n_I1_vec),
    
    # Severity (mean only)
    severity_I0_mean = mean(severity_I0_vec),
    severity_I1_mean = mean(severity_I1_vec),
    
    # Shannon
    shannon_I0_mean = mean(shannon_I0_vec),
    shannon_I0_sd = sd(shannon_I0_vec),
    shannon_I1_mean = mean(shannon_I1_vec),
    shannon_I1_sd = sd(shannon_I1_vec),
    delta_shannon_mean = mean(delta_shannon),
    delta_shannon_ci_lower = ci_shannon["lower"],
    delta_shannon_ci_upper = ci_shannon["upper"],
    
    # Hill q=0
    hill_q0_I0_mean = mean(hill_q0_I0_vec),
    hill_q0_I0_sd = sd(hill_q0_I0_vec),
    hill_q0_I1_mean = mean(hill_q0_I1_vec),
    hill_q0_I1_sd = sd(hill_q0_I1_vec),
    delta_hill_q0_mean = mean(delta_hill_q0),
    delta_hill_q0_ci_lower = ci_hill_q0["lower"],
    delta_hill_q0_ci_upper = ci_hill_q0["upper"],
    
    # Hill q=1
    hill_q1_I0_mean = mean(hill_q1_I0_vec),
    hill_q1_I0_sd = sd(hill_q1_I0_vec),
    hill_q1_I1_mean = mean(hill_q1_I1_vec),
    hill_q1_I1_sd = sd(hill_q1_I1_vec),
    delta_hill_q1_mean = mean(delta_hill_q1),
    delta_hill_q1_ci_lower = ci_hill_q1["lower"],
    delta_hill_q1_ci_upper = ci_hill_q1["upper"],
    
    # Hill q=2
    hill_q2_I0_mean = mean(hill_q2_I0_vec),
    hill_q2_I0_sd = sd(hill_q2_I0_vec),
    hill_q2_I1_mean = mean(hill_q2_I1_vec),
    hill_q2_I1_sd = sd(hill_q2_I1_vec),
    delta_hill_q2_mean = mean(delta_hill_q2),
    delta_hill_q2_ci_lower = ci_hill_q2["lower"],
    delta_hill_q2_ci_upper = ci_hill_q2["upper"],
    
    stringsAsFactors = FALSE
  ))
}

cat("\n\n=== ALL SCENARIOS COMPLETED ===\n")
```

## Export Raw Results to CSV

```{r export_csv, echo=TRUE}
# Save raw numeric results to CSV (for archiving and potential re-analysis)
write.csv(results_block1, "Additive_Block1_Results_Raw.csv", row.names = FALSE)
cat("Raw results saved to 'Additive_Block1_Results_Raw.csv'\n")
```

## Create Publication-Ready Tables

```{r create_tables, echo=TRUE}
# =============================================================================
# MAIN MANUSCRIPT TABLE
# Shows: p(L), p(I), c, Δ Shannon [95% CI], Δ Hill q=1 [95% CI]
# =============================================================================

main_table_df <- results_block1 %>%
  transmute(
    `p(L)` = p_L,
    `p(I)` = p_I,
    `c` = cor,
    `Δ Shannon [95% CI]` = sprintf("%.2f [%.2f, %.2f]", 
                                    delta_shannon_mean, 
                                    delta_shannon_ci_lower, 
                                    delta_shannon_ci_upper),
    `Δ Hill q=1 [95% CI]` = sprintf("%.2f [%.2f, %.2f]", 
                                     delta_hill_q1_mean, 
                                     delta_hill_q1_ci_lower, 
                                     delta_hill_q1_ci_upper)
  )

# Create flextable for main manuscript
ft_main <- flextable(main_table_df) %>%
  theme_booktabs() %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 4, part = "all") %>%
  set_caption("Additive Model - Block 1")

# Display main table
ft_main
```

```{r create_supplementary_table, echo=TRUE}
# =============================================================================
# SUPPLEMENTARY TABLE
# Shows all columns with:
# - Mean only for sample size and severity
# - Mean ± SD for Shannon and Hill values
# - Mean [95% CI] for deltas
# =============================================================================

supp_table_df <- results_block1 %>%
  transmute(
    `p(L)` = p_L,
    `p(I)` = p_I,
    `c` = cor,
    
    # Shannon (mean ± SD for groups, mean [95% CI] for delta)
    `Shannon (I=0)` = sprintf("%.2f ± %.2f", shannon_I0_mean, shannon_I0_sd),
    `Shannon (I=1)` = sprintf("%.2f ± %.2f", shannon_I1_mean, shannon_I1_sd),
    `Δ Shannon` = sprintf("%.2f [%.2f, %.2f]", 
                          delta_shannon_mean, 
                          delta_shannon_ci_lower, 
                          delta_shannon_ci_upper),
    
    # Hill q=0 (mean ± SD for groups, mean [95% CI] for delta)
    `Hill q=0 (I=0)` = sprintf("%.2f ± %.2f", hill_q0_I0_mean, hill_q0_I0_sd),
    `Hill q=0 (I=1)` = sprintf("%.2f ± %.2f", hill_q0_I1_mean, hill_q0_I1_sd),
    `Δ Hill q=0` = sprintf("%.2f [%.2f, %.2f]", 
                           delta_hill_q0_mean, 
                           delta_hill_q0_ci_lower, 
                           delta_hill_q0_ci_upper),
    
    # Hill q=1 (mean ± SD for groups, mean [95% CI] for delta)
    `Hill q=1 (I=0)` = sprintf("%.2f ± %.2f", hill_q1_I0_mean, hill_q1_I0_sd),
    `Hill q=1 (I=1)` = sprintf("%.2f ± %.2f", hill_q1_I1_mean, hill_q1_I1_sd),
    `Δ Hill q=1` = sprintf("%.2f [%.2f, %.2f]", 
                           delta_hill_q1_mean, 
                           delta_hill_q1_ci_lower, 
                           delta_hill_q1_ci_upper),
    
    # Hill q=2 (mean ± SD for groups, mean [95% CI] for delta)
    `Hill q=2 (I=0)` = sprintf("%.2f ± %.2f", hill_q2_I0_mean, hill_q2_I0_sd),
    `Hill q=2 (I=1)` = sprintf("%.2f ± %.2f", hill_q2_I1_mean, hill_q2_I1_sd),
    `Δ Hill q=2` = sprintf("%.2f [%.2f, %.2f]", 
                           delta_hill_q2_mean, 
                           delta_hill_q2_ci_lower, 
                           delta_hill_q2_ci_upper)
  )

# Create flextable for supplementary material
ft_supp <- flextable(supp_table_df) %>%
  theme_booktabs() %>%
  fontsize(size = 7, part = "all") %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  padding(padding = 2, part = "all") %>%
  autofit() %>%
  set_caption("Supplementary Table: Additive Model - Block 1: Detailed Heterogeneity Results")

# Display supplementary table
ft_supp
```

## Export Tables to Word

```{r export_tables, echo=TRUE}
# Export main manuscript table
doc_main <- read_docx() %>%
  body_add_par("Table X. Additive Model - Block 1: Sensitivity Analysis", style = "heading 1") %>%
  body_add_par("") %>%
  body_add_flextable(ft_main) %>%
  body_add_par("") %>%
  body_add_par("Note: Δ = difference between groups (I=1 minus I=0). Values represent mean [95% CI] across 1000 simulations.", style = "Normal")

print(doc_main, target = "Table_Additive_Block1_Main.docx")
cat("Main table saved to 'Table_Additive_Block1_Main.docx'\n")

# Export supplementary table
doc_supp <- read_docx() %>%
  body_add_par("Supplementary Table X. Additive Model - Block 1: Detailed Results", style = "heading 1") %>%
  body_add_par("") %>%
  body_add_flextable(ft_supp) %>%
  body_add_par("") %>%
  body_add_par("Note: Group values shown as mean ± SD. Δ values shown as mean [95% CI]. All statistics calculated across 1000 simulations.", style = "Normal")

print(doc_supp, target = "Table_Additive_Block1_Supplementary.docx")
cat("Supplementary table saved to 'Table_Additive_Block1_Supplementary.docx'\n")
```
